The role of advection versus diffusion in intraventricular transport

### 2025-03-03 paper/main.tex
Request citations in Introduction.

\section{Introduction}
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid (CSF). CSF is a
clear liquid produced by the choroid plexus and the fluid flows
through the ventricular system and subarachnoid space.  By
transporting growth factors, nutrients, neurotransmitters and
metabolites and removing waste products, CSF regulates brain
development and homeostasis.  Disturbance of the CSF circulation can
lead to brain disorders, such as hydrocephalus where accumulation of
CSF often requires emergency surgical treatment.

Motion of CSF occurs through multiple physiological processes,
including the beating motion of motile cilia lining the ventricular walls,
CSF secretion by the choroid plexus, and pressure gradients generated
by the cardiac cycle, respiration, and bodily movement. Motile cilia
are microscopic hair-like structures that beat in a metachronal wave
pattern to mediate fluid flow. Due to their microscopic size, ranging
from 6 to 15 microns, the cilia contribute to flow primarily at the surface
of the brain ventricles. In contrast, fluid flow induced by the 
cardiac and respiratory cycles account for the bulk flow in the brain
ventricles~\cite{Vinje2019RespiratoryMeasurements},
as well as the CSF flow in the perivascular spaces.

To date, it remains poorly understood how cilia contribute to the
overall fluid flow and solute transport within the brain.
These questions have been challenging to tackle due to technical
limitations. Notably, cilia are microscopic and beat at high
frequencies ranging from 10 to 30 Hz, while other fluid dynamic
processes in the brain are slower and span larger volumes, making
cilia dynamics challenging to monitor in vivo. Until now, most studies have
analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles; however, they failed to identify the role of pressure
gradients associated with respiration and cardiac pulsations in CSF flow
and the accompanying transport. Similarly, while MRI technology can measure CSF movement
in humans, it cannot detect the microscopic movements of cilia to
understand their impact in the large human brain. Consequently, we
lack a comprehensive understanding of how cilia contribute to CSF
motion and solute transport, and whether cilia dysfunction will
alter CSF composition and lead to brain disorders. This is particularly
important since ventricular enlargement, known as hydrocephalus,
is observed in animal models and humans with motile cilia deficiency,
indicating that the cilia play a crucial role for healthy brain functioning.

The technical limitations of studying cilia-mediated flow and transport opens the
possibility of mathematical modeling and computational investigations. Numerous previous studies
have focused on modeling the interaction between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction},
and cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.

In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
First, we computed flow patterns and simulated solute transport with a computational model. To
this end, we used the geometry of the zebrafish larval brain ventricles,
for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies. Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages,
while still being well-conserved with mammals. 
Next, we validated our computation model with experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as the ventricular geometry, greatly
impact solute distribution. Altogether, our work presents a computational
framework together with new concepts of how cilia contribute to distribution of molecules and
particles within the brain ventricles.

>>>>

\section{Introduction}

% I1: What is the problem - from big picture to focus
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{CITATTION}. CSF is a clear liquid produced by the choroid
plexus that flows through the ventricular system and subarachnoid
space. By transporting growth factors, nutrients, neurotransmitters
and metabolites as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{CITATTION}. Disturbance of the
CSF circulation can lead to brain disorders, such as hydrocephalus
where accumulation of CSF often requires emergency surgical
treatment~\cite{CITATTION}.

% MER: Add this to the paragraph above.
%
% This is particularly important since ventricular enlargement, known
% as hydrocephalus, is observed in animal models and humans with
% motile cilia deficiency, indicating that the cilia play a crucial
% role for healthy brain functioning.

% I2: What do we know about the problem?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia lining the ventricular walls, CSF secretion by the choroid
plexus, and pulsatile pressure gradients generated by the cardiac
cycle, respiration, neural activity, and bodily
movement~\cite{CITATION}. Motile cilia are microscopic hair-like
structures that beat in a metachronal wave pattern at frequencies of
10--30 Hz~\cite{CITATION}. Given their microscopic size, ranging from
6 to 15 microns, the cilia clearly contributes to flow near the
ventricular surface~\cite{CITATION}. On the other hand, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at scale in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
  causemann2025insilico, fultz2019coupled}.

%How the cilia contribute to overall CSF flow and solute transport
%remains poorly understood however.

%These questions have been challenging to tackle due to technical
%limitations.

% I3: What do we not know about the problem and what do we want to
% know?

%As a result, cilia dynamics are technically challenging to monitor
%in-vivo.
%
%Until now, most studies have analyzed one physical mechanism at a
%time, especially in mammals.
%
\mer{MER: Let's look at this next paragraph together. It should clearly lay out what we do not know about the problem (and what we want to find out). Be precise. Refer to specific studies and what specifically we (do not) know. What we know belongs nicely in the paragraph above.}
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles; however, they failed to identify the role of pressure
gradients associated with respiration and cardiac pulsations in CSF
flow and the accompanying transport. Similarly, while MRI technology
can measure CSF movement in humans, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain. Consequently, we lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport, and whether cilia
dysfunction will alter CSF composition and lead to brain
disorders. The technical limitations of studying cilia-mediated flow and transport opens the
possibility of mathematical modeling and computational investigations. Numerous previous studies
have focused on modeling the interaction between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction},
and cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
First, we computed flow patterns and simulated solute transport with a computational model. To
this end, we used the geometry of the zebrafish larval brain ventricles,
for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies. Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages,
while still being well-conserved with mammals. 
Next, we validated our computation model with experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as the ventricular geometry, greatly
impact solute distribution. Altogether, our work presents a computational
framework together with new concepts of how cilia contribute to distribution of molecules and
particles within the brain ventricles.
